---
title: "IGP CORRELATION CODES"
author: "PROJECT TEAM 1"
date: "August 22"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---
```{r}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE,comment=NA)
```

```{r}

employment_data <- read.csv("merged_data.csv")
```

#glimpse the loaded csv file to check structure of dataset such as no of rows, no of column and variables data type


```{r}
source("mypackages.R")
source("helperFunctions.R")
library(dplyr)

```

```{r}
library(tidyr)
```


```{r}
filter_date<- subset(employment_data,employment_data$Date == c("2019","2020","2021"))
```

```{r}
science <- filter_date[,c("Date","Country","X21.Science..research..engineering.and.technology.professionals")]
```

Here we make use of pivot wider which converts long data into wide data using the function pivot wider.
```{r}
science1 <- science%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X21.Science..research..engineering.and.technology.professionals'
  )
```

```{r}
with(science1, cor(England, Scotland))
with(science1, cor.test(England, Scotland))
```
#This signifies a high positive correlation between the two countries.science, tech and engineering professions experienced a shift in the upward direction for both countries.


```{r}
construction <- filter_date[,c("Date","Country","X53.Skilled.construction.and.building.trades")]
```

```{r}
construction1 <- construction%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X53.Skilled.construction.and.building.trades'
  )
```

```{r}
with(construction1, cor(Scotland, Wales))
with(construction1, cor.test(Scotland, Wales))
```

```{r}
transport <- filter_date[,c("Date","Country","X82.Transport.and.mobile.machine.drivers.and.operatives")]
```

```{r}
transport1 <- transport%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X82.Transport.and.mobile.machine.drivers.and.operatives'
  )
```

```{r}
with(transport1, cor(Scotland, Wales))
with(transport1, cor.test(Scotland, Wales))
```
#This signifies a high positive correlation between the two countries.Transport professions experienced a shift in the upward direction for both countries and there's a 6.6% probability that the null hypothesis is correct.


```{r}
protect <- filter_date[,c("Date","Country","X33.Protective.service.occupations")]
```

```{r}
protect1 <- protect%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X33.Protective.service.occupations'
  )
```


```{r}
with(protect1, cor(Wales, Northern_Ireland))
with(protect1, cor.test(Wales, Northern_Ireland))
```

```{r}
health <- filter_date[,c("Date","Country","X22.Health.professionals")]
```

```{r}
health1 <- health%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X22.Health.professionals'
  )
```

```{r}
with(health1, cor(England, Wales))
with(health1, cor.test(England, Wales))
```
```{r}
care <- filter_date[,c("Date","Country","X61.Caring.personal.service.occupations")]
```

```{r}
care1 <- care%>%
  pivot_wider(
   names_from = 'Country',
   values_from = 'X61.Caring.personal.service.occupations'
  )
```

```{r}
with(care1, cor(Scotland, England))
with(care1, cor.test(Scotland, England))
```
