---
title: "IGP DATA ANALYSIS"
author: "PROJECT TEAM 1"
date: "August 22"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE,comment=NA) 
```
 Data Description:
 This dataset is part of a larger dataset that has been collected to help to know the impact of Covid 19 Pandemic on the employment profession in the 4 regions of the UK and to also observe the various shifts that occurred in different professions as a result of this, paying particular attention to professions that gained, loss and constant.

```{r}
source("mypackages.R")
source("helperFunctions.R")

library(tidyr)

```

```{r}
# load dataset
employment_data <- read.csv("merged_data.csv")

#glimpse the loaded csv file to check structure of dataset such as no of rows, no of column and variables data type
```

```{r}
library(dplyr)

```
Data Preparation

We are interested in modelling the employment by professions where Country is equal to Scotland.

 Rows of data with this property are selected and used to perform the rest of the analysis.


```{r}
#Select rows that satisfies the condition specified in the modelling
employ_scotland<-subset(employment_data,employment_data$Country == "Scotland")
```

```{r}
tinytex::install_tinytex()
```


```{r}
glimpse(employ_scotland)
```
The following exploratory data analysis will be done using descriptive statistics, exploratory graphs and linear correlation.

summary(): To measure central tendency by checking for variation in mean, median, also for outliers. If variation or difference between mean and median is much then it means there's outliers.

Skimr: this is another method used for summarizing data. It gives a broad set of summary statistics with a single function call.

```{r}
summary(employ_scotland)
```
```{r}
is.null(employ_scotland)
skimr::skim(employ_scotland)
```
Summary shows there are no outliers as variations between mean and median in all professions are minimal.
We checked for null values and got a boolean FALSE.

EXPLORATORY GRAPHS

5 professions were selected and grouped into 3 categories and all visually analysed.

```{r}
category1 <- employ_scotland[,c("Date","Country","X11.Corporate.managers.and.directors","X21.Science..research..engineering.and.technology.professionals","X22.Health.professionals","X23.Teaching.and.educational.professionals","X24.Business..media.and.public.service.professionals")]
```

Here we make use of pivot longer which converts wide data into long data using the function pivot longer.Line graphs were plotted showing 5 professions in each categories.
```{r}
#library(tidyverse)
category1_long <- category1%>%
  pivot_longer(
   cols = starts_with("X")
  )
```

```{r}
ggplot(category1_long,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="EMPLOYMENTS CATEGORY1 IN SCOTLAND")
```


```{r}
category2 <- employ_scotland[,c("Date","Country","X33.Protective.service.occupations", "X34.Culture..media.and.sports.occupations","X41.Administrative.occupations","X42.Secretarial.and.related.occupations","X51.Skilled.agricultural.and.related.trades")]
```

```{r}
category2_long <- category2%>%
  pivot_longer(
   cols = starts_with("X")
  )
```

```{r}
ggplot(category2_long,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="EMPLOYMENTS CATEGORY2 IN SCOTLAND")
```
```{r}
category3 <- employ_scotland[,c("Date","Country","X53.Skilled.construction.and.building.trades", "X61.Caring.personal.service.occupations", "X62.Leisure..travel.and.related.personal.service.occupations","X81.Process..plant.and.machine.operatives","X82.Transport.and.mobile.machine.drivers.and.operatives")]
```

```{r}
category3_long <- category3%>%
  pivot_longer(
   cols = starts_with("X")
  )
```

```{r}
ggplot(category3_long,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="EMPLOYMENTS CATEGORY3 IN SCOTLAND")
```

```{r}
protective<- employ_scotland[,c("Date","Country","X33.Protective.service.occupations")]
```

```{r}
protective2 <- protective%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

From the categories plotted, professions with notable changes were plotted individually to further highlight the changes.

```{r}
ggplot(protective2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="PROTECTIVE PROFESSION IN SCOTLAND")
```
This shows there was a continual decline with effect initializing from 2019 but with significant COVID effect in 2020



```{r}
science<- employ_scotland[,c("Date","Country","X21.Science..research..engineering.and.technology.professionals")]
```

```{r}
science2 <- science%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

```{r}
ggplot(science2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="SCIENCE,TECHNOLOGY AND ENGINEERING PROFESSION IN SCOTLAND")
```
This shows there was a continual increase with effect initializing from 2019 but with significant COVID effect in 2020.
```{r}
admin<- employ_scotland [,c("Date","Country","X41.Administrative.occupations")]
```

```{r}
admin2 <- admin%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

```{r}
ggplot(admin2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="ADMINISTRATIVE PROFESSION IN SCOTLAND")
```
This shows there was a slight increase in 2019 but with significant COVID effect increase in 2020
```{r}
construct<- employ_scotland[,c("Date","Country","X53.Skilled.construction.and.building.trades")]
```

```{r}
construct2 <- construct%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```


```{r}
ggplot(construct2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="CONSTRUCTION AND BUILDING TRADES PROFESSION IN SCOTLAND")
```
This shows there was a continual decline with effect initializing from 2019 but with significant COVID effect decrease in 2020.

```{r}
agric<- employ_scotland[,c("Date","Country","X51.Skilled.agricultural.and.related.trades")]

```

```{r}
agric2 <- agric%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

```{r}
ggplot(agric2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="AGRICULTURAL PROFESSION IN SCOTLAND")
```
This shows there was a continual decline with effect initializing from 2019 but with significant COVID effect decrease in 2020.
```{r}
caring <- employ_scotland[,c("Date","Country","X61.Caring.personal.service.occupations")]
```

```{r}
caring2 <- caring%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

```{r}
ggplot(caring2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="CARING PROFESSION IN SCOTLAND")
```
This shows there was a decline with effect initializing from 2018 but with a slight change in trend in 2020.

```{r}
corporate <- employ_scotland[,c("Date","Country","X11.Corporate.managers.and.directors")]

```

```{r}
corporate2 <- corporate%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```

```{r}
ggplot(corporate2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="CORPORATE PROFESSION IN SCOTLAND")

```
This shows there was a  decline with effect initializing from 2018 but with a slight change in trend in 2020


```{r}
health <- employ_scotland[,c("Date","Country","X22.Health.professionals")]

```

```{r}
health2 <- health%>%
  pivot_longer(
   cols = starts_with("X") 
  )
```


```{r}
ggplot(health2,aes(x=Date,y=value,col=as.factor(name)))+geom_line()+xlab("years")+ylab("number of employed per profession")+theme_minimal()+labs(title="HEALTH PROFESSION IN SCOTLAND")
```
This shows there was a decline with effect initializing from 2018 but with almost no significance in 2020.

